---
title: "Bayes Factor"
author: "Richèl J.C. Bilderbeek"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bayes Factor}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Goal

Measure the Bayes factor between JC69 and GTR.

```{r}
library(mcbette)
library(testthat)
```

Set a random seed:

```{r}
set.seed(314)
```

BEAST2 must be installed:

```{r}
if (!beastier::is_beast2_installed()) {
  message("BEAST2 must be installed. Tip: use 'beastier::install_beast2()'")
}
```

The BEAST2 package `NS` must be installed:

```{r}
if (!mauricer::is_beast2_ns_pkg_installed()) {
  message(
    "The BEAST2 'NS' package must be installed. ",
    "Tip: use 'mauricer::install_beast2_pkg(\"NS\")'"
  )
}
```

Create data frame:

```{r}
df <- data.frame(
  site_model_name = c("JC69", "GTR"),
  marg_log_lik = c(-1234.567, -23456.78),
  weight = c(0.9, 0.1)
)
```

Interpretation of the Bayes factor [1]:

```{r}
interpret_bayes_factor <- function(bayes_factor) {
  if (length(bayes_factor) != 1 || !is.numeric(bayes_factor)) {
    stop("'bayes_factor' must be one numeric value")
  }

  if (bayes_factor < 10^0.0) {
    "in favor of other model"
  } else if (bayes_factor < 10^0.5) {
    "barely worth mentioning"
  } else if (bayes_factor < 10^1.0) {
    "substantial"
  } else if (bayes_factor < 10^1.5) {
    "strong"
  } else if (bayes_factor < 10^2.0) {
    "very strong"
  } else {
    "decisive"
  }
}
expect_equal(interpret_bayes_factor(0.5), "in favor of other model")
expect_equal(interpret_bayes_factor(1.5), "barely worth mentioning")
expect_equal(interpret_bayes_factor(8.5), "substantial")
expect_equal(interpret_bayes_factor(12.5), "strong")
expect_equal(interpret_bayes_factor(85.0), "very strong")
expect_equal(interpret_bayes_factor(123.0), "decisive")

expect_error(
  interpret_bayes_factor(c(0.1, 0.2)), 
  "'bayes_factor' must be one numeric value"
)
expect_error(
  interpret_bayes_factor("nonsense"), 
  "'bayes_factor' must be one numeric value"
)
expect_error(
  interpret_bayes_factor(NA), 
  "'bayes_factor' must be one numeric value"
)
expect_error(
  interpret_bayes_factor(NULL), 
  "'bayes_factor' must be one numeric value"
)
```

## Overview

 * Create a data set that *is* JC69
 * Calculate the mar
 * Compare these posteriors using the Bayes factor
 * Interpret the Bayes factor

## Methods

Create an articifical PBD phylogeny with a long time to speciate. That is,
a tree with all speciation event observed early in time (close to the root):
 
```{r fig.width=7}
phylogeny <- ape::read.tree(text = "((((A:12, B:12):1,C:13):1,D:14):1, E:15);")
crown_age <- beautier::get_crown_age(phylogeny)
ape::plot.phylo(phylogeny)
```

Create JC69 alignment and save it to a FASTA file:

```{r fig.width=7}
sequence_length <- 1000

alignment_phydat <- phangorn::simSeq(
  phylogeny,
  l = sequence_length,
  rate = 1.0 / crown_age,
  rootseq = rep(c('a', 'c', 'g', 't'), each = sequence_length / 4)
)
testit::assert(class(alignment_phydat) == "phyDat")
alignment_dnabin <- ape::as.DNAbin(alignment_phydat)

fasta_filename <- tempfile(pattern = "bayes_factor_", fileext = ".fasta")
phangorn::write.phyDat(
  alignment_dnabin,
  file = fasta_filename,
  format = "fasta"
)
image(alignment_dnabin)
```

From that alignment, get the marginal likelihood of the two desired models:


```{r}
if (mauricer::is_beast2_ns_pkg_installed()) {
  df <- mcbette::est_marg_liks(
    fasta_filename = fasta_filename, 
    site_models = list(
      beautier::create_jc69_site_model(), 
      beautier::create_gtr_site_model()
    ),
    clock_models = list(beautier::create_strict_clock_model()),
    tree_priors = list(beautier::create_bd_tree_prior()),
    epsilon = 100.0
  )
  knitr::kable(df)
}
```

The Bayes factor (usually denoted as K) can be used for model comparison. 
It can be calculated from the marginal likelihoods ratio:

```{r}
marg_lik_jc <- exp(Rmpfr::mpfr(df$marg_log_lik[1], 512)) 
marg_lik_gtr <- exp(Rmpfr::mpfr(df$marg_log_lik[2], 512))
bayes_factor <- marg_lik_jc / marg_lik_gtr
interpretation <- interpret_bayes_factor(as.numeric(bayes_factor))
paste(interpretation, "in favor of JC69")
```

There is also another way to compare models: by comparing their relative weights:

```{r}
site_model_name <- NULL; rm(site_model_name) # nolint, fixes warning: no visible binding for global variable
weight <- NULL; rm(weight) # nolint, fixes warning: no visible binding for global variable

ggplot2::ggplot(
  df, ggplot2::aes(x = site_model_name, y = weight)
) + ggplot2::geom_col()
```

## References

 * [1] H. Jeffreys (1961). The Theory of Probability (3rd ed.). Oxford. p. 432
